---
title: 'DataViz: Dynamic Visualization with Gapminder'
author: "Heather Kitada Smalley"

output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Motivating Example: Gapminder
#### Fertility vs Life Expectancy

In this demo we are going to learn about dynamic graphics by using the Gapminder data on Fertility.  These data are discussed in the Hans Rosling TED Talk (start at 2:30) <https://youtu.be/hVimVzgtD6w?t=149>.

For more detail on how the data were wrangled please look at the article "How to build Animated Charts like Hans Rosling — doing it all in R" by Tristan Ganry <https://towardsdatascience.com/how-to-build-animated-charts-like-hans-rosling-doing-it-all-in-r-570efc6ba382>.

### 1) Start Simple: `ggplot`

```{r message=FALSE, warning=FALSE}
library(tidyverse)

fert <- read_csv("https://raw.githubusercontent.com/kitadasmalley/FA2020_DataViz/main/data/gapminderFert.csv")

head(fert)

p <- fert %>%
  filter(year == 2015) %>%
  ggplot(aes(x = fert, y = life, size = pop, color = continent)) +
  labs(x = "Fertility Rate", y = "Life expectancy at birth (years)",
       caption = "(Based on data from Hans Rosling - gapminder.com)",
       color = 'Continent',size = "Population (millions)") +
  ylim(30,100) +
  geom_point()

p
```

### 2) Add some interaction: `plotly`

The `plotly` package can be used to created interactive graphics.  Although it is an entirely different paradigm of programming graphics, it does have a very nice function, `ggplotly` that can be added on to ggplot objects.

#### Use the `tooltip` to learn more about each point

```{r message=FALSE, warning=FALSE}
#install.packages("plotly")
library(plotly)

p <- fert %>%
  ggplot(aes(x = fert, y = life, size = pop, color = continent, frame = year)) +
  labs(x = "Fertility Rate", y = "Life expectancy at birth (years)",
       caption = "(Based on data from Hans Rosling - gapminder.com)",
       color = 'Continent', size = "Population (millions)") +
  ylim(30,100) +
  geom_point()

ggplotly(p)
```

### 3) Now let's animate! `gganimate`
```{r message=FALSE, warning=FALSE}
#install.packages("gganimate")
library(gganimate)
#install.packages("gifski")
library(gifski)

p1 <- fert %>%
  ggplot(aes(x = fert, y = life, size = pop, color = continent, frame = year)) +
  labs(x = "Fertility Rate", y = "Life expectancy at birth (years)",
       caption = "(Based on data from Hans Rosling - gapminder.com)",
       color = 'Continent', size = "Population (millions)") +
  ylim(30,100) +
  geom_point() +
  ggtitle("Year: {round(frame_time, 0)}") +
  transition_time(year) +
  ease_aes("linear") +
  enter_fade() +
  exit_fade()

animate(p1, fps = 4, width = 600, height = 400, renderer = gifski_renderer())
anim_save("output.gif")
```

Notice that this animation is a little choppy, because there is one observation every year.  You can use the `tweener` package to interpolate and make smoother transitions <https://www.r-bloggers.com/2017/05/create-smooth-animations-in-r-with-the-tweenr-package/>.
