---
title: 'DataViz: Interactive Demos'
author: "Heather Kitada Smalley"

output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following demos are modified from <https://www.r-graph-gallery.com/interactive-charts.html>.

### I. Examples with `plotly`

#### 1) Scatterplots

```{r warning=FALSE, message=FALSE}
fert <- read.csv("https://raw.githubusercontent.com/kitadasmalley/FA2020_DataViz/main/data/gapminderFert.csv",
               header = TRUE)

library(tidyverse)
#install.packages("plotly")
library(plotly)

p <- fert %>%
  ggplot(aes(x=fert, y=life, size = pop, color = continent, frame = year)) +
  labs(x="Fertility Rate", y = "Life expectancy at birth (years)",
       caption = "(Based on data from Hans Rosling - gapminder.com)",
       color = 'Continent', size = "Population (millions)") +
  ylim(30,100) +
  geom_point(aes(text = Country))

ggplotly(p)
```

#### 2) Area Charts and Time Series

```{r warning=FALSE, message=FALSE}
# Load dataset from github
data <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/3_TwoNumOrdered.csv", header=T)
data$date <- as.Date(data$date)

# Usual area chart
p <- data %>%
  ggplot( aes(x=date, y=value)) +
  geom_area(fill="#69b3a2", alpha=0.5) +
  geom_line(color="#69b3a2") +
  ylab("bitcoin price ($)")

# Turn it interactive with ggplotly
p <- ggplotly(p)
p
```

### II. Example with `heatmaply`

Here is an example showing 8 general features like population or life expectancy for about 30 countries in 2015. Data come from the French National Institute of Demographic Studies <https://www.ined.fr/en/everything_about_population/data/all-countries/?lst_continent=908&lst_pays=926>

Heatmap (left to right):
* Pop
* Birth rate
* Mortality rate
* Life expectancy
* Infant Mortality
* Children per woman
* Growth rate
* Population aged 65

```{r warning=FALSE, message=FALSE}
library(viridis)

#install.packages("heatmaply")
library(heatmaply)

# Load data
data <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/multivariate.csv", header = T, sep = ";")
head(data)
colnames(data) <- gsub("\\.", " ", colnames(data))

# Select a few country
data <- data %>%
  filter(Country %in% c("France", "Sweden", "Italy", "Spain", "England", "Portugal", "Greece", "Peru", "Chile", "Brazil", "Argentina", "Bolivia", "Venezuela", "Australia", "New Zealand", "Fiji", "China", "India", "Thailand", "Afghanistan", "Bangladesh", "United States of America", "Canada", "Burundi", "Angola", "Kenya", "Togo")) %>%
  arrange(Country) %>%
  mutate(Country = factor(Country, Country))

# Matrix format
mat <- data
rownames(mat) <- mat[,1]
mat <- mat %>% dplyr::select(-Country, -Group, -Continent)
mat <- as.matrix(mat)

head(mat)

# Heatmap
p <- heatmaply(mat,
               dendrogram = "none",
               xlab = "", ylab = "",
               main = "",
               scale = "column",
               margins = c(60,100,40,20),
               grid_color = "white",
               grid_width = 0.00001,
               titleX = FALSE,
               hide_colorbar = TRUE,
               branches_lwd = 0.1,
               label_names = c("Country", "Feature:", "Value"),
               fontsize_row = 5, fontsize_col = 5,
               labCol = colnames(mat),
               labRow = rownames(mat),
               heatmap_layers = theme(axis.line=element_blank())
)
p
```

### III. `streamgraph`

Here is a great resource about the streamgraph htmlwidgtet R Package! <https://hrbrmstr.github.io/streamgraph/>

Using the `babynames` dataset from (1880-2013)
```{r}
library(dplyr)
library(babynames)

#install.packages("remotes")
#remotes::install_github("hrbrmstr/streamgraph")

library(streamgraph)

babynames %>%
  filter(grepl("^Kr", name)) %>%
  group_by(year, name) %>%
  tally(wt=n) %>%
  streamgraph("name", "n", "year")
```

Or you can add a drop down

```{r}
babynames %>%
  filter(grepl("^H", name)) %>%
  group_by(year, name) %>%
  tally(wt=n) %>%
  streamgraph("name", "n", "year", offset="zero", interpolate="linear") %>%
  sg_legend(show=TRUE, label="H- names: ")

```


### IV. Chord diagrams with `chorddiag`

```{r}
#install.packages("devtools")
#devtools::install_github("mattflor/chorddiag")
library(chorddiag)

# Create dummy data
m <- matrix(c(11975,  5871, 8916, 2868,
              1951, 10048, 2060, 6171,
              8010, 16145, 8090, 8045,
              1013,   990,  940, 6907),
            byrow = TRUE,
            nrow = 4, ncol = 4)

# A vector of 4 colors for 4 groups
haircolors <- c("black", "blonde", "brown", "red")
dimnames(m) <- list(have = haircolors,
                    prefer = haircolors)
groupColors <- c("#000000", "#FFDD89", "#957244", "#F26223")

# Build the chord diagram:
p <- chorddiag(m, groupColors = groupColors, groupnamePadding = 50)
p
```


### V. Network Diagrams with `networkD3`

```{r}
#install.packages("networkD3")

library(networkD3)

# create a dataset:
data <- data.frame(
  from=c("A", "A", "B", "D", "C", "D", "E", "B", "C", "D", "K", "A", "M"),
  to=c("B", "E", "F", "A", "C", "A", "B", "Z", "A", "C", "A", "B", "K")
)


# Plot
p <- simpleNetwork(data, height="100px", width="100px",
                   Source = 1,                 # column number of source
                   Target = 2,                 # column number of target
                   linkDistance = 10,          # distance between node. Increase this value to have more space between nodes
                   charge = -900,              # numeric value indicating either the strength of the node repulsion (negative value) or attraction (positive value)
                   fontSize = 14,              # size of the node names
                   fontFamily = "serif",       # font og node names
                   linkColour = "#666",        # colour of edges, MUST be a common colour for the whole graph
                   nodeColour = "#69b3a2",     # colour of nodes, MUST be a common colour for the whole graph
                   opacity = 0.9,              # opacity of nodes. 0=transparent. 1=no transparency
                   zoom = T                    # Can you zoom on the figure?
)

p

```
