---
title: "Exploration and Experimentation with ggplot2 submission"
author: "Rob Osborne"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---

## Install Required Packages
```{r, warning=FALSE, message=FALSE}
# Set CRAN mirror
options(repos = c(CRAN = "https://cran.rstudio.com/"))

# Install packages
if (!requireNamespace("ggplot2movies", quietly = TRUE)) {
  install.packages("ggplot2movies")
}
if (!requireNamespace("ggthemes", quietly = TRUE)) {
  install.packages("ggthemes")
}
if (!requireNamespace("tinytex", quietly = TRUE)) {
  install.packages("tinytex")
}

# Install TinyTeX if not already installed
if (!tinytex::is_tinytex()) {
  tinytex::install_tinytex()
}
```


## Load Required Packages
```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(ggplot2movies)
library(ggthemes)
library(tinytex)
library(dplyr)
library(scales)
```


## Load Data
```{r, warning=FALSE, message=FALSE}
data(movies) # Load the dataset
head(movies) # Preview the dataset
# ?movies
```


## Plot 1: Ratings by Genre

#### Data Preparation
```{r, warning=FALSE, message=FALSE}
# Create a new column 'genre' in the 'movies' data frame.
# This column categorizes each movie based on its genre.
movies$genre <- ifelse(movies$Action == 1, "Action",
                       ifelse(movies$Comedy == 1, "Comedy", "Other"))
```


### 1st Draft

Limit the plotting by genre to only two genres: action and comedy.

```
ggplot(movies, aes(x = genre, y = rating)) +
  geom_boxplot() +
  labs(
    title = "Ratings by Genre",
    x = "Genre",
    y = "Rating"
  ) +
  theme_economist()
```

### 2nd Draft

Update the plot to use the violin geom for a different view.

```
ggplot(movies, aes(x = genre, y = rating, fill = genre)) +
  geom_violin() +
  labs(
    title = "Ratings by Genre",
    x = "Genre",
    y = "Rating"
  ) +
  theme_void()
```

### Final Draft

Expand the genres. All genres should include (action, animation, comedy, doumentary, drama, romance, short, and other). This plot will show the distribution of ratings for each genre with a dashed line across the x axis for the overall move average rating.

```{r, warning=FALSE, message=FALSE}
movies$genre <- ifelse(movies$Action == 1, "Action",
ifelse(movies$Animation == 1, "Animation",
ifelse(movies$Comedy == 1, "Comedy",
ifelse(movies$Documentary == 1, "Documentary",
ifelse(movies$Drama == 1, "Drama",
ifelse(movies$Romance == 1, "Romance",
ifelse(movies$Short == 1, "Short",
"Other")))))))

average_rating <- mean(movies$rating, na.rm = TRUE)

ggplot(movies, aes(x = genre, y = rating, fill = genre)) +
  geom_violin() +
  geom_hline(yintercept = average_rating, linetype = "dashed", color = "red") +
  labs(
    title = "Ratings by Genre",
    x = "Genre",
    y = "Rating"
  ) +
  theme_minimal() +
  theme(legend.position = "top", panel.grid.major.x = element_blank())
```


## Plot 2: Budget vs Ratings

#### Data Preparation

```{r, warning=FALSE, message=FALSE}
# Remove rows where the 'budget' column has NA values
movies_clean <- movies[!is.na(movies$budget), ]
```

### 1st Draft

Plot a basic scatter plot of rating vs. budget. This first plot is pretty messy as most values are plotted on the left side of the graph making it difficult to understand.

```
ggplot(movies, aes(x = budget, y = rating)) +
  geom_point(alpha = 0.5) +
  labs(title = "Movie Rating vs. Budget",
       x = "Budget",
       y = "Rating") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

### 2nd Draft

Shifting to a log scale for the x-axis to better visualize the data as before most of the data was plotted on the left of the graph making it difficult to understand.
Using the `scales` package to format the x-axis labels with commas.

```
ggplot(movies_clean, aes(x = budget, y = rating)) +
  geom_point(alpha = 0.5) +
  geom_jitter() +
  scale_x_log10(labels = comma) +
  labs(title = "Movie Rating vs. Budget (NA Removed)",
       x = "Budget",
       y = "Rating") +
  theme_hc() + scale_colour_hc()
```

### Final Draft

For the final draft we are going to use a 2D bin plot to better visualize where most of the data is. This will help us understand the distribution of the data better. The data shows that most movies have a budget between $1 - 100 million and a rating between 5 and 7.

```{r, warning=FALSE, message=FALSE}
ggplot(movies_clean, aes(x = budget, y = rating)) +
  stat_bin_2d() +
  scale_x_log10(labels = comma) +
  labs(title = "Movie Rating vs. Budget (NA Removed)",
       x = "Budget",
       y = "Rating") +
  theme_classic() +
  theme(legend.position = "top")
```


## Plot 3: Year vs Length

#### Data Preparation
```{r, warning=FALSE, message=FALSE}
# Store the average movie length per year in a new data frame
avg_length_per_year <- movies %>%
  group_by(year) %>%
  summarize(avg_length = mean(length, na.rm = TRUE))
```

### 1st Draft

Plot a bar chart of movie length vs. year.

```
ggplot(movies, aes(x = year, y = length)) +
  geom_bar(stat = "identity") +
  labs(title = "Movie Length vs. Year",
       x = "Year",
       y = "Length") +
  theme_minimal()
```

### 2nd Draft

Calculate the average movie length per year and plot it as a bar chart. This will give a better sense of how movie lengths (based on an average) have changed over the years.

```
ggplot(avg_length_per_year, aes(x = year, y = avg_length)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Movie Length vs. Year",
       x = "Year",
       y = "Average Length") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```

### Final Draft

Calculate the average movie length per year and plot it as a line chart with a smoothed line to show the trend more clearly. The trend shows us that the average movie length have been increasing from 1890 to 1980. This is likely due to the advancement of filming technology.

```{r, warning=FALSE, message=FALSE}
ggplot(avg_length_per_year, aes(x = year, y = avg_length)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  labs(title = "Average Movie Length vs. Year",
       x = "Year",
       y = "Average Length") +
  theme_minimal()
```


## Plot 4: Genre Popularity Over Time

#### Data Preparation
```{r, warning=FALSE, message=FALSE}
# Create a new column 'genre' in the 'movies' data frame.
# This column categorizes each movie based on its genre.
movies$genre <- ifelse(movies$Action == 1, "Action",
ifelse(movies$Animation == 1, "Animation",
ifelse(movies$Comedy == 1, "Comedy",
ifelse(movies$Documentary == 1, "Documentary",
ifelse(movies$Drama == 1, "Drama",
ifelse(movies$Romance == 1, "Romance",
ifelse(movies$Short == 1, "Short",
"Other")))))))

# Calculate the number of movies per genre per year
genre_counts <- movies %>%
  group_by(year, genre) %>%
  summarize(count = n())
```

### 1st Draft

Calculate the number of movies per genre per year and plot the number of movies per genre over time.

```
ggplot(genre_counts, aes(x = year, y = count, color = genre)) +
  geom_bar(stat = "identity") +
  labs(title = "Genre Popularity Over Time",
       x = "Year",
       y = "Number of Movies") +
  theme_minimal()
```

### 2nd Draft

Change the plot to use a dodged bar chart to better compare the number of movies per genre over time.

```
ggplot(genre_counts, aes(x = year, y = count, fill = genre)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Genre Popularity Over Time",
       x = "Year",
       y = "Number of Movies",
       fill = "Genre") +
  theme_minimal()
```

### Final Draft

Calculate the number of movies per genre per year and plot the number of movies per genre over time. This plot shows us that between years 1980 and 2010 there was an increase in popularity for comedy movies.

```{r, warning=FALSE, message=FALSE}
ggplot(genre_counts, aes(x = year, y = count, color = genre)) +
  geom_line(size = 1) +
  labs(title = "Genre Popularity Over Time",
       x = "Year",
       y = "Number of Movies") +
  theme_minimal() +
  theme(legend.position = "top")
```
