---
title: 'Introduction to ggplot2: Points'
author: "Rob Osborne"
output: html_document
---


In this lab we will be learning about some basics with `ggplot2`.

## Part I: Learning about your data

### Step 1: Load in the data

First let's load in the `diamonds` dataset.  This data set is in the `tidyverse` package, so make sure that that library is called first.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
data("diamonds")
```

### Step 2: Learn a little bit about this data

The `str` function allows you to learn about the structure of a dataset.

```{r}
str(diamonds)
```

### It's Your Turn! Learning by doing

1. How many rows are in `diamonds`?  How many columns?
    53940 row and 10 columns

2. Which variables in `diamonds` are categorical? Which variables are continuous?
(Hint: look at the output for `str()`)
    Categorical: cut, color, clarity
    Continuous: carat, depth, table, price, x, y, z

3. What does the `table` variable describe?  Read the help for `?diamonds` to find out.
    `table` is the width of top of diamond relative to widest point (43-95)

## Part II: Start with a basic scatterplot

Here is a simple scatterplot of `price` vs `carat`:

```{r warning=FALSE, message=FALSE}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point()
```

What do you observe?

### It's Your Turn! Learning by doing

1. Run `ggplot(data=diamonds)`.  What do you see?
    A blank canvas
2. Make a scatterplot of `price` vs `depth`.
```{r}
ggplot(data = diamonds, aes(x = depth, y = price)) +
  geom_point()
```

3. What happens if you make a scatterplot of `cut` vs `clarity`? Why is the plot not useful?
```{r}
ggplot(data = diamonds, aes(x = cut, y = clarity)) +
  geom_point()
```
It doesn't say much. The data is categorical and the plot is treating it as continuous.


## Part III: Aesthetic Mappings

Aesthetic mappings translate

### A) Color

#### Color Gradient for Ordinal Data
```{r}
# If using a categorical variable each category will have a color
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) + geom_point()
```

#### Unique colors for Nominal Data
```{r}
# if not ordered..
ggplot(data = diamonds, aes(x = carat, y = price, color = as.character(clarity))
) + geom_point()
```

#### Saturation Gradient for Numeric
```{r}
# If using a numeric variable there will be a color gradient
ggplot(data = diamonds, aes(x = carat, y = price, color = depth)) + geom_point()
```

You can also apply a single color to all the data points by specifying the `color` outside of the aesthetic mapping.

```{r}
ggplot(data = diamonds, aes(x = carat, y = price, color = "orange")) +
  geom_point()
```

### B) Transparency
Great for show overlapping points

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = .4)
```


### C) Shape
Only useful for categorical variables if you have a small number of categories
```{r}
ggplot(data = diamonds, aes(x = carat, y = price, shape = cut)) +
  geom_point()
```


### D) Size
Should be mapped to a numeric variable
```{r}
ggplot(data = diamonds, aes(x = carat, y = price, size = depth)) +
  geom_point()
```


### It's Your Turn! Learning by doing

1. What’s gone wrong with this code? Why are the points not blue?
The color is not being mapped to the variable correctly. The color should specify the color parameter outside the aes() function.

```{r}
ggplot(diamonds, aes(x = carat, y = price, size = depth, shape = cut, color = as.character(clarity))) +
  geom_point(color = "blue")
```

2. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?
The aesthetics


```{r}
ggplot(diamonds, aes(x = carat, y = price, size = depth, shape = cut, color = as.character(clarity))) +
  geom_point()
```

3. What happens if you map the same variable to multiple aesthetics?
```{r}
ggplot(data = diamonds, aes(x = carat, y = price, shape = clarity, color = clarity)) + geom_point()
```

4. What happens if you map an aesthetic to something other than a variable name, like `aes(colour = carat < 3)`? Note, you’ll also need to specify x and y.
```{r}
ggplot(data = diamonds, aes(x = carat, y = price, color = carat < 3)) +
  geom_point()
```

## Part IV: Facets

Sometimes it's useful to look at subgroups within our data.  We can do this with facets.

### `facet_wrap()`

You can specify a single discrete variable to facet by and R organize plots to fill the space.

```{r}
## INSERT YOUR CODE HERE ##
```

### `facet_grid()`

You can also create a grid of graphs.  The first argument to the function specifies rows and the second columns.

```{r}
## Grid
## INSERT YOUR CODE HERE ##
```

If you prefer to not facet in the rows or columns dimension, use a . instead of a variable name, e.g. `+ facet_grid(. ~ color)`.

### It's Your Turn! Learning by doing

1. What happens if you facet on a continuous variable?

2. What plots does the following code make? What does `.` do?

```{r eval=FALSE}
ggplot(diamonds, aes(carat, price))+
  geom_point()+
  facet_grid(color~.)

ggplot(diamonds, aes(carat, price))+
  geom_point()+
  facet_grid(.~cut)
```

3. When using `facet_grid()` you should usually put the variable with more unique levels in the columns. Why?
