---
title: "DataViz: Alternative to Pie Charts"
author: "INSERT NAME HERE"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Commuting Data

The following data show commuting transportation methods for states on the West Coast of the United States.

```{r warning=FALSE, message=FALSE}
library(tidyverse)

#rds is an special r dataframe
#commute <- readRDS("~/Desktop/DataViz Reading/Week 5 Data and Code/commute.rds")

# csv format
commute<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/DATA502/FALL2024/Week5/commute.csv", header=TRUE)

# a subset of states for class
states <- c("or", "ak", "ca", "wa", "id", "nv")
commute_nw <- filter(commute, state %in% states)

```

## Pie Chart

Pie charts are common, but not great for:

  1) Many categories
  2) Comparison

```{r}
ggplot(commute_nw) +
  geom_bar(aes(x = "", y = prop, fill = factor(transport_type)),
    stat = "identity", width = 1) +
  facet_wrap(~ state_name) +
  coord_polar(theta = "y") +
  theme_minimal(18) +
  xlab("") + ylab("") +
  theme(axis.text = element_blank()) +
  scale_fill_discrete(name = "Transporation") +
  labs(title = "People in Oregon commute by bicycle more \nthan in other (nearby) states")
```

## Activity

### Step 1: Sketch an alternative display of the data that would be an improvement.

  * Graphs should employ perceptual tasks as high on the ranking as possible.
  * Keep important comparisons close.

### Step 2:  Share your sketch with your group

  * Discuss what you think was most effective and what was not
  * Choose one idea to implement together in `ggplot2`



```{r}
# filter data based on Bicyle
bicyle_commute_nw <- filter(commute_nw, transport_type == "Bicycle")

ggplot(bicyle_commute_nw) +
  geom_bar(aes(x = reorder(state_name, -prop), y = prop, fill = state == "or"),
    stat = "identity", width = 0.7) +
  theme_minimal() +
  xlab("State") + ylab("Proportion") +
  scale_fill_discrete(name = "Transporation") +
  labs(title = "People in Oregon commute by bicycle more \nthan in other (nearby) states")
```

```{r echo=FALSE}
# filter data based on Bicyle
bicyle_commute_nw <- filter(commute_nw, transport_type == "Bicycle")

ggplot(bicyle_commute_nw) +
  geom_boxplot(aes(x = state_name, y = prop)) +
  theme_minimal() +
  xlab("State") + ylab("Proportion") +
  scale_fill_discrete(name = "Transportation") +
  labs(title = "Distribution of Bicycle Commute Proportions by State")
```

### Step 3:  Share with class and submit
