---
title: 'DataViz: Ensemble Graphics'
author: "Heather Kitada Smalley"

output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Ensemble Graphics

These examples modified from *Chapter 12 Ensemble Graphics and Case Studies in Graphical Data Analysis with R,by Antony Unwin*

### Motivating Example 1: Coffee

```{r}
#install.packages("pgmm")
library(pgmm)
data("coffee")
head(coffee)

names(coffee)
```

#### 1.1 Exploration:

##### Is that any missing data?

A unicorn!
```{r}
#install.packages("naniar")
library(naniar)

## LOOK FOR MISSING DATA

vis_miss(coffee) # Sees what data is missing. Included in naniar package

```

#### 1.1.1 Visualize One Categorical Variable: Bar Charts

##### What is the distribution of coffee varieties?
```{r message=FALSE, warning=FALSE}
library(tidyverse)

# barcharts for categorical vars

ggplot(coffee, aes(factor(Variety))) +
  geom_bar()
```

##### What is the distribution of coffee origin countries?
```{r}
length(unique(coffee$Country))
# 29 different countries (with odd spellings)

ggplot(coffee, aes(Country)) +
  geom_bar() +
  coord_flip()
```


#### 1.1.2 Visualize Multiple Categorical Variables: Mosaic plots

##### What is the distribution of coffee varieties across country?

```{r}
#mosaic for categorica
#install.packages("vcd")
library(vcd)

## MOSAIC PLOT

mosaic(Variety ~ Country, data = coffee)

```

Here's another example of a mosaic plot that is built into this package:
```{r}
#example
mosaic(HairEyeColor, shade = TRUE, legend = TRUE)
```

#### 1.1.3 Visualize Continuous Variables: Standardized boxplots

Look for outliers

```{r}
# distribution of continuous variables
# standardized!

boxplot(scale(coffee[, 3:14]))
```


#### 1.1.4 Visualize Relationships between Two Continuous Variables: Pairs plots

```{r}
# scatterplot matrix of continuous
pairs(coffee[, 3:14], pch = 16)
```

```{r}
# scatterplot matrix of continuous
## ONLY LOOK AT VARIABLES 3 through 7

pairs(coffee[, 3:7], pch = 16)
```

```{r}
#install.packages("GGally")
library(GGally)

## ONLY LOOK AT VARIABLES 3 through 7

ggpairs(coffee[, 3:7])
```

#### 1.2 Polishing/Explanation

* Labels
* Names of Varieties
* Names of Variables
* Consistent colors

##### 1.2.1 Names of Varieties

```{r}
coffee <- within(coffee, Type <- ifelse(Variety == 1, "Arabica", "Robusta"))
head(coffee)
```

##### 1.2.2 Names of Variables
```{r}

## WHAT DOES THE FUNCTION abbreviate do?

names(coffee) <- abbreviate(names(coffee), 8)
names(coffee)

```

#### 1.3 Graphics

##### 1.3.1  Bargraph
```{r}

## MAKE PLOT 1 (save it as "a")
a <- ggplot(coffee, aes(x = Type)) +
  geom_bar(aes(fill = Type)) +
  scale_fill_manual(values = c("grey70", "red")) +
  guides(fill = FALSE) +
  ylab("")

a
```

##### 1.3.2 Scatterplot
```{r}
## MAKE PLOT 2 (save it as "b")
b <- ggplot(coffee, aes(x = Fat, y = Caffine, colour = Type)) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("grey70", "red"))

b
```

##### 1.3.3 Parallel Coordinate Plot
```{r}
#install.packages("GGally")
library(GGally)

## MAKE PLOT 3 (save it as "c")
## A PARALELL COORDINATE PLOT

c <- ggparcoord(coffee[order(coffee$Type), ], columns = 3:14,
              groupColumn = "Type", scale = "uniminmax",
              mapping = aes(size = 1)) +
  xlab("") + ylab("") +
  theme(legend.position = "none") +
  scale_colour_manual(values = c("grey", "red")) +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank())
c


```

#### 1.4 Ensemble Graphic

```{r}
#install.packages("gridExtra")
library(gridExtra)

## COMBINE WITH GRID ARRANGE

grid.arrange(arrangeGrob(a, b, ncol = 2, widths = c(1, 2)), c, nrow = 2)
```

### Case Study: Fertility

Mothers in this dataset had two or more children.
```{r message=FALSE, warning=FALSE}
### Fertility dataset
#install.packages("AER")
library(AER)
data("Fertility")

head(Fertility)
```


#### 2.1 Exploration

##### 2.1.1 What is the distribution of age?

"As expected, the numbers of women with two or more children rise with age, although it is perhaps surprising that the pattern is so regular before dropping off for the two oldest cohorts"

```{r}
## GEOM BAR
## INSERT CODE HERE ###
```

##### 2.1.2 Is the distribution of age different across demographic groups?

The same pattern can be seen across all race groups.

```{r}
##FACET WRAP

## INSERT CODE HERE ###
```

##### 2.1.3 Does whether or not the mother worked affect number of children?

In these data, a majority of the women either didn't work or worked, with some in-between.

We can look at the proportions:

```{r}
with(Fertility, prop.table(table(work)))
```

#### 2.2 Graphics

Since we are using the same dataset and similar geometries we can use the same "base" and then add modifiers to it.

##### 2.2.1 Age and Weeks Worked
```{r}
p0<-ggplot(Fertility)+geom_bar(binwidth=1)+ylab("")
p1<- p0+aes(x=age)
p2<-p0+aes(x=work)+xlab("Weeks worked in 1979")

p1
p2
```

##### 2.2.2 Other Variables
```{r}
k<-ggplot(Fertility)+geom_bar()+ylab("")+ylim(0, 250000)
p3<-k+aes(x=morekids) + xlab("has more children")
p4<-k+aes(x=gender1) + xlab("first child")
p5<-k+aes(x=gender2) + xlab("second child")
p6<-k+aes(x=afam) + xlab("African American")
p7<-k+aes(x=hispanic) + xlab("Hispanic")
p8<-k+aes(x=other) + xlab("other race")
```

##### 2.2.3 Now combine!
```{r}
grid.arrange(arrangeGrob(p1, p2, ncol=2, widths = c(3,3)),
             arrangeGrob(p3, p4, p5, p6, p7, p8, ncol=6),
             nrow=2, heights=c(1.25, 1))
```

#### 2.3 Spinogram plots

Note that these are different than colored bar charts because the widths also show the proportions

##### 2.3.1 More than two kids by age
```{r}
doubledecker(morekids~age, data=Fertility,
             gp=gpar(fill=c("grey90", "forestgreen")),
             spacing=spacing_equal(0))
```

##### 2.3.2 More than two kids by genders of previous children
```{r}
doubledecker(morekids~gender1+gender2, data=Fertility,
             gp=gpar(fill=c("grey90", "forestgreen")))
```

##### 2.3.3 More than two kids by age and genders of previous children
```{r}
doubledecker(morekids~age+gender1+gender2, data=Fertility,
             gp=gpar(fill=c("grey90", "forestgreen")),
             spacing=spacing_dimequal(c(0.1, 0, 0, 0)))
```

### 3. Case Studies

#### 3.1 Moral statistics of France
```{r}
#install.packages("HistData")
library(HistData)
data(Guerry)
head(Guerry)

```

#### 3.2 Airbags and car accidents
```{r}
#install.packages("DAAG")
library(DAAG)
data(nassCDS)
head(nassCDS)
```

#### 3.3 Athletes' blood measurements
```{r}
library(DAAG)
data(ais)
head(ais)

```

#### 3.4 Marijuana arrests
```{r}
#install.packages("effects")
library(effects)
data("Arrests")
head(Arrests)
```

#### 3.5 Crohn's disease
```{r}
#install.packages("robustbase")
library(robustbase)
data("CrohnD")
head(CrohnD)
```

#### 3.6 Footballers in the four major European leagues
```{r}
#install.packages("SportsAnalytics")
library(SportsAnalytics)
data(EURO4PlayerSkillsSep11)
head(EURO4PlayerSkillsSep11)
```

#### 3.7 Decathlon
```{r}
#install.packages("GDAdata")
library(GDAdata)
data("Decathlon")
head(Decathlon)
```
