---
title: 'DataViz: Scales Lab'
author: "INSERT NAME HERE"
output: html_document
---

*Adapted from a lab written by Prof Charlotte Wickham*

# LAB DIRECTIONS:

## Exploring flight delays
To experiment with some features of scales we are going to work with a heatmap of the proportion of flights that have a departure delay of more that 15 minutes (`prop_over_15`) at the George Bush Intercontinental Airport (IAH) by day of the week (`DayOfWeek`) and departure hour (`DepHour`).

### Step 1: Get the data
These data come from the `hflights` package, but some summarization is done for you.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
iah <- read_csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/main/DATA502/FA2023/R_Markdown/Week8/iah_flightSummary.csv")

str(iah)
```

### Step 2: Make a heatmap:

```{r}
# make sure days of week are displayed in the right order
iah$DayOfWeek <- factor(iah$DayOfWeek,
  levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))

p <- ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15))

p
```

### Step 3: Experiment with `scale_fill_xxx`

We are interested in changing the scale for the fill of the tiles so we should be adding a `scale_fill_xxx function`, where `xxx` is replaced by a scale name. (If you are working with points and have mapped color to a variable, you would use `scale_color_xxx`.) When the fill aesthetic is mapped to a continuous variable, the default is `scale_fill_gradient`.

##### Verify that adding this explicitly to the plot doesn’t change anything:

```{r}
p + scale_fill_gradient2()
```

##### Change end point colors:

We can change the colors at the endpoints of the scale by specifying the `high` and `low` arguments, for example,

```{r}
p + scale_fill_gradient("Proportion", high = "white", low = "springgreen4")
```

#### **Experiment!**

Use different named colors for high and low arguments to scale_fill_gradient. You can see a list of the named colors with `colors()`, or a useful .pdf is at: <http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf>

### Step 4: Find better color scales

It’s surprisingly hard using scale_fill_gradient to find “nice” color gradients, and there is no guarantee they’ll correspond to any perceptually sound palettes like those you’ve seen in the readings.

```{r}
# install.packages("colorspace")
library(colorspace)
```

colorspace has a whole set of named palettes, you can see them all with:

```{r}
hcl_palettes(plot = TRUE)
```

Notice they are broken into sections for qualitative, sequential and diverging palettes.

To apply a colorspace palette to a ggplot2 plot, you use the a function that takes the form:

```{}
scale_<aesthetic>_<datatype>_<colorscale>()
```

Where `<datatype>` is one of `discrete` or `continuous`, and `<colorscale>` is one of `qualitative`, `sequential`, or `diverging`, and you’ll pass the palette name as the argument.

So, for instance to add the `"Mint"` palette, you could use:

```{r}
p + scale_fill_continuous_qualitative("Cold")
```


I’ve used `fill` because our original plot mapped `prop_over_15` to `fill`, `continuous` because `prop_over_15` is continuous, and `sequential` because `"Mint"` is a sequential scale.

#### **Experiment!**

Try 3-4 other colorspace sequential scales.

```{r}
### EXPERIMENT HERE ###
# p + scale_fill_binned_sequential("Mint")
# p + scale_fill_binned_sequential("PinkYl")
# p + scale_fill_continuous_sequential("Plasma")
# p + scale_fill_binned_sequential("ag_Sunset")
```


You could use a diverging scale, but since colorspace assumes zero is the midpoint you’ll only get one half of the palette:

```{r}
p +  scale_fill_continuous_diverging("Blue-Red")
```

### Step 5: Transform data and change limits

Let’s look at the average departure delay of delayed flights

```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = avg_delay_delayed)) +
  scale_fill_gradient()

```

*Why is this plot so uninformative?*

Try limiting the fill scale to only show values between 0 and 120:

```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = avg_delay_delayed)) +
  scale_fill_gradient(limits = c(0, 120))
```

Instead of using `limits` try using `trans = "log10"`. **What happens?**

Compare that to mapping fill to the log transformed average delay of delayed flights:

```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = log10(avg_delay_delayed))) +
  scale_fill_gradient()
```

Other transformations are listed in the “See Also” section of `?trans_new`.

An alternative approach to deal with a few very large numbers is to turn the continuous variable into a discrete one by binning it:

```{r}
iah <- iah %>%
  mutate(avg_delay_cut = cut(avg_delay, breaks = c(-5, 0, 15, 30, 60, 1000)))
```

Then you’ll need to use the discrete form of the scale:
```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = avg_delay_cut))  +
  scale_fill_discrete_sequential("Plasma")
```

#### **Experiment!**

1. Try adding a different discrete sequential color palette.
1. Try adding a discrete qualitative scale. Is the one you tried useful for this variable?
1. What happens if you add a discrete diverging scale? Why isn’t this useful in this example?

```{r}
### SPACE TO EXPERIMENT ###
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = avg_delay_cut))  +
  scale_fill_discrete_sequential("Mint")
```


### Step 6: Simulating color vision deficiency

Experiment Save one of your plots so far as a PNG file, upload it at <http://hclwizard.org/cvdemulator/>. and click on the “All” tab to see it as it may appear to someone with each of the different color vision deficiencies. **Are the colors still distinguishable?**

### More information on Guides

Guides control how legends appear. Take a look at <http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/>

and the examples at <https://ggplot2.tidyverse.org/reference/guide_colourbar.html>

and <https://ggplot2.tidyverse.org/reference/guide_legend.html>

to see what can be changed.