---
title: 'Data Viz: Recreating Shot Plot'
author: "INSERT NAME"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Los Angeles Lakers Shooting Heatmap

In this exercise we will recreate a graphic that describes positions on the basketball court the average points scored from that position and the frequency at which shot are taken from this position.

### The Data

These data represent the LA Lakers 2009/2010 season and come from http://www.basketballgeek.com.  We are going to access this data from my github.
```{r}
shots_sum <- read.csv("https://raw.githubusercontent.com/kitadasmalley/DATA502/main/FALL2021/Data/shots_sum.csv",
                      header = TRUE)
head(shots_sum)

```

## Step 1: Start with a simple scatterplot
```{r message=FALSE, warning=FALSE}
library(tidyverse)

# Step 1 : Scatterplot
ggplot(shots_sum, aes(x, y)) + geom_point()
```

## Step 2: Add color
```{r}
# Step 2: Add color
ggplot(shots_sum, aes(x, y)) +
  geom_point(aes(color = avg_points))
```

## Step 3: Size the bubbles
```{r}
# Step 3: Size bubbles
ggplot(shots_sum, aes(x, y)) +
  geom_point(aes(color = avg_points, size = num_shots))
```

#### What is going on with the big dot in the center?

Its messing up the sizing!

Let's explore:
```{r}
subset(shots_sum, num_shots > 3000)
```

## Step 4: Remove point and re-plot
```{r}
# Step 4: Subset the data to remove the "outlier"
# shots_sub <- subset(shots_sum, num_shots > 3000)
# dim(shots_sub)

# this can also be done with filer
shots_sub <- shots_sum %>% filter(num_shots <= 3000)
dim(shots_sub)

# graph again
ggplot(shots_sub, aes(x, y)) +
  geom_point(aes(color = avg_points, size = num_shots))
```

## Step 5: Change limits to focus within the 3 point boundary
```{r}
ggplot(shots_sub, aes(x, y)) +
  geom_point(aes(color = avg_points, size = num_shots)) +
  ylim(0, 35)
```

## Step 6: X and Y distance should be in same distance
```{r}
# Step 6: x and y distance should be in same distance
ggplot(shots_sub, aes(x, y)) +
  geom_point(aes(color = avg_points, size = num_shots)) +
  ylim(0, 35) +
  coord_equal()
```

## Step 7: Change color from green to red
```{r}
# Step 7: Change color from green to red
ggplot(shots_sub, aes(x, y)) +
  geom_point(aes(color = avg_points, size = num_shots)) +
  scale_color_distiller(palette = "RdYlGn") +
  ylim(0, 35) +
  coord_equal()
```

## Step 8: Change size scale
```{r}
# Step 8: Change size scale
ggplot(shots_sub, aes(x, y)) +
  geom_point(aes(color = avg_points, size = num_shots)) +
  scale_color_distiller(palette = "RdYlGn") +
  scale_size(trans = "sqrt", range = c(0.1, 5)) +
  ylim(0, 35) +
  coord_equal()
```

## Step 9: Polishing
```{r}
# Step 9: Polishing
ggplot(shots_sub, aes(x, y, color = avg_points, size = num_shots)) +
  geom_point() +
  scale_color_distiller("Points", palette = "RdYlGn") +
  scale_size(trans = "sqrt", range = c(0.1, 5)) +
  ylim(0, 35) +
  coord_equal() +
  theme_classic() +
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
```
