---
title: 'DataViz: Polishing'
author: "INSERT NAME HERE"

output: html_document
---
*Adapted from a lab written by Prof Charlotte Wickham*



### The data
These data come from the `hflights` package, but some summarization is done for you.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(colorspace)
iah <- read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/main/DATA502/FA2023/R_Markdown/Week8/iah_flightSummary.csv")

str(iah)
```

### Start with this heatmap:

The heatmap represents the proportion of flights that have a departure delay of more that 15 minutes (`prop_over_15`) at the George Bush Intercontinental Airport (IAH) by day of the week (`DayOfWeek`) and departure hour (`DepHour`).

```{r}
# make sure days of week are displayed in the right order
iah$DayOfWeek <- factor(iah$DayOfWeek,
  levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))

p <- ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15))

p
```

### Step 1 : `scale_x_continuous`

#### Breaks

```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15), colour = "grey50") +
  scale_x_continuous("Departure time", breaks = c(0, 6, 12, 18, 24))
```

#### Labels

```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15), colour = "grey50") +
  scale_x_continuous("Departure time",
    breaks = c(0, 6, 12, 18, 24),
    labels = c("midnight", "6am", "noon", "6pm", "midnight"))
```
#### Limits

0 and midnight are the same thing!

```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15), colour = "grey50") +
  scale_x_continuous("Departure time",
    breaks = c(0, 6, 12, 18, 24),
    labels = c("midnight", "6am", "noon", "6pm", "midnight"),
    limits = c(0, 23))
```

#### Expansion (Padding Space)

```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15), colour = "grey50") +
  scale_x_continuous("Departure time",
    breaks = c(0, 6, 12, 18, 24),
    labels = c("midnight", "6am", "noon", "6pm", "midnight"),
    expand = c(0, 0),
    limits = c(0, 23))
```



### Step 2: `scale_y_discrete`

#### Label

```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15), colour = "grey50") +
  scale_x_continuous("Departure time",
    breaks = c(0, 6, 12, 18, 24),
    labels = c("midnight", "6am", "noon", "6pm", "midnight"),
    expand = c(0, 0),
    limits = c(0, 23)) +
  scale_y_discrete("Departure day")
```

### Step 3:  `scale_fill_continuous_sequential`

We're going to need the color space package for this.

#### Palette
```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15), colour = "grey50") +
  scale_x_continuous("Departure time",
    breaks = c(0, 6, 12, 18, 24),
    labels = c("midnight", "6am", "noon", "6pm", "midnight"),
    expand = c(0, 0),
    limits = c(0, 23)) +
  scale_y_discrete("Departure day") +
  scale_fill_continuous_sequential(
    name = "Flights delayed \nmore than 15 mins",
    palette = "YlOrRd")
```

#### Breaks

```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15), colour = "grey50") +
  scale_x_continuous("Departure time",
    breaks = c(0, 6, 12, 18, 24),
    labels = c("midnight", "6am", "noon", "6pm", "midnight"),
    expand = c(0, 0),
    limits = c(0, 23)) +
  scale_y_discrete("Departure day") +
  scale_fill_continuous_sequential(
    name = "Flights delayed \nmore than 15 mins",
    palette = "YlOrRd",
    breaks = c(0, .25, 0.5, .75, 1))
```

#### Labels
```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15), colour = "grey50") +
  scale_x_continuous("Departure time",
    breaks = c(0, 6, 12, 18, 24),
    labels = c("midnight", "6am", "noon", "6pm", "midnight"),
    expand = c(0, 0),
    limits = c(0, 23)) +
  scale_y_discrete("Departure day") +
  scale_fill_continuous_sequential(
    name = "Flights delayed \nmore than 15 mins",
    palette = "YlOrRd",
    breaks = c(0, .25, 0.5, .75, 1),
    labels = c("0%", "25%", "50%", "75%", "100%"))
```

#### Guide (Legend)

```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15), colour = "grey50") +
  scale_x_continuous("Departure time",
    breaks = c(0, 6, 12, 18, 24),
    labels = c("midnight", "6am", "noon", "6pm", "midnight"),
    expand = c(0, 0),
    limits = c(0, 23)) +
  scale_y_discrete("Departure day") +
  scale_fill_continuous_sequential(
    name = "Flights delayed \nmore than 15 mins",
    palette = "YlOrRd",
    breaks = c(0, .25, 0.5, .75, 1),
    labels = c("0%", "25%", "50%", "75%", "100%"),
    expand = c(0, 0),
    guide = "colorbar")
```

### Step 4: Title and subtitle

```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15), colour = "grey50") +
  scale_x_continuous("Departure time",
    breaks = c(0, 6, 12, 18, 24),
    labels = c("midnight", "6am", "noon", "6pm", "midnight"),
    expand = c(0, 0),
    limits = c(0, 23)) +
  scale_y_discrete("Departure day") +
  scale_fill_continuous_sequential(
    name = "Flights delayed \nmore than 15 mins",
    palette = "YlOrRd",
    breaks = c(0, .25, 0.5, .75, 1),
    labels = c("0%", "25%", "50%", "75%", "100%"),
    expand = c(0, 0),
    guide = "colorbar") +
  labs(title = "Take an early flight from IAH to avoid delays",
    subtitle = "Based on all departing flights from George Bush Intercontinental Airport (IAH) in 2011")
```



### Step 5: Theme

```{r}
ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15), colour = "grey50") +
  scale_x_continuous("Departure time",
    breaks = c(0, 6, 12, 18, 24),
    labels = c("midnight", "6am", "noon", "6pm", "midnight"),
    expand = c(0, 0),
    limits = c(0, 23)) +
  scale_y_discrete("Departure day") +
  scale_fill_continuous_sequential(
    name = "Flights delayed \nmore than 15 mins",
    palette = "YlOrRd",
    breaks = c(0, .25, 0.5, .75, 1),
    labels = c("0%", "25%", "50%", "75%", "100%"),
    expand = c(0, 0),
    guide = "colorbar") +
  labs(title = "Take an early flight from IAH to avoid delays",
    subtitle = "Based on all departing flights from George Bush Intercontinental Airport (IAH) in 2011") +
  theme_classic() +
  theme(axis.ticks.y = element_blank(), axis.line.y = element_blank()) + coord_equal()
```



### Step 6: Order Days (Relevel)
```{r}
iah$DayOfWeek <- factor(iah$DayOfWeek,
                        levels = c("Sun", "Sat", "Fri", "Thu", "Wed", "Tue","Mon"))

ggplot(iah, aes(DepHour, DayOfWeek)) +
  geom_tile(aes(fill = prop_over_15), colour = "grey50") +
  scale_x_continuous("Departure time",
    breaks = c(0, 6, 12, 18, 24),
    labels = c("midnight", "6am", "noon", "6pm", "midnight"),
    expand = c(0, 0),
    limits = c(0, 23)) +
  scale_y_discrete("Departure day") +
  scale_fill_continuous_sequential(
    name = "Flights delayed \nmore than 15 mins",
    palette = "YlOrRd",
    breaks = c(0, .25, 0.5, .75, 1),
    labels = c("0%", "25%", "50%", "75%", "100%"),
    expand = c(0, 0),
    guide = "colorbar") +
  labs(title = "Take an early flight from IAH to avoid delays",
    subtitle = "Based on all departing flights from George Bush Intercontinental Airport (IAH) in 2011") +
  theme_classic() +
  theme(axis.ticks.y = element_blank(), axis.line.y = element_blank()) + coord_equal()
```


### Step 7: `ggsave()`

```{r}
#ggsave("polishedDelays.png", height = 3.5, width = 10)
```