---
title: 'Data Viz: Introduction to ggplot2'
author: "INSERT NAME HERE"
output:
  html_document:
    toc: true
    collapsed: false
    toc_float: true
    toc_depth: 5
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Part I: Review `geom_point`

### 0. Basics

Recall our work from last class with using `geom_point`:

#### Load example data
```{r warning = FALSE, message = FALSE}
library(tidyverse)
data("diamonds")
str(diamonds)
```

#### Starting with a simple scatterplot:

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_point()
```

### 1. Aesthetic Mapping

#### A. Color

##### Categorical vs Continuous Data
```{r}
# If using a categorical variable each category will have a color
ggplot(diamonds, aes(carat, price, color = clarity)) +
  geom_point()

# if not ordered..
ggplot(diamonds, aes(carat, price, color = as.character(clarity))) +
  geom_point()

# If using a numeric variable there will be a color gradient
ggplot(diamonds, aes(carat, price, color = depth)) +
  geom_point()
```

##### CAUTION

Be careful of where you place the color statement
```{r}
ggplot(diamonds, aes(carat, price, color = "blue")) +
  geom_point()

ggplot(diamonds, aes(carat, price)) +
  geom_point(color = "blue")
```

#### B. Transparency
```{r}
ggplot(diamonds, aes(carat, price, alpha = clarity)) +
  geom_point()
```


#### C. Shape
```{r}
ggplot(diamonds, aes(carat, price, shape = clarity)) +
  geom_point()
```


#### D. Size
```{r}
ggplot(diamonds, aes(carat, price, size = clarity)) +
  geom_point()
```

### 2. Facet

Use facets to explore subsets
```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_point() +
  facet_grid(. ~ cut)
#   facet_grid(cut ~ .) # flip the grid. Good for x comparisons
#   facet_grid(clarity ~ cut) # See comparisons between variables
#   facet_wrap(~ cut) # wrap the grid
```

## Part II: Geometries

### 1. Using a single geometry

#### A. Scatterplot
We have already constructed scatterplots with `geom_point()` let's take a look at other geometries.

#### B. Smooth
```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_smooth()
#   geom_smooth(se = FALSE) # remove standard error
#   geom_smooth(method = "lm") # linear model
```

#### C. Boxplot
```{r}
# Vanilla
# ggplot(diamonds, aes(carat, price)) +
#   geom_boxplot()

# Side-by-side
ggplot(diamonds, aes(y = price, fill = cut)) +
  geom_boxplot()
```

#### D. Bar Chart
```{r}
ggplot(diamonds, aes(x = cut)) +
  geom_bar()
```

#### E. Time Series / Line Plots

Here we will take a quick detour from the diamonds dataset to look at data from IMDB on action movies:
```{r}
# install.packages("ggplot2movies")
library(ggplot2movies)
data(movies)

movies_yg <- group_by(movies, year, Action)
movies_sum <- summarise(movies_yg, n_movies = n())

## INSERT YOUR CODE HERE ##
```

### 2 Using multiple geometries at the same time

#### Point and Smooth

Adding `geom_smooth` on top of your `geom_point` can help you see what relationships/trends may exist between you response and explanatory variables.

```{r}
# POINT AND SMOOTH
## INSERT YOUR CODE HERE ##
```

#### Refining your `geom_smooth`

##### 1. Removing the Standard Error Band

```{r}
# TURNING OFF SE (STANDARD ERROR BARS) IN GEOM_SMOOTH
## INSERT YOUR CODE HERE ##
```

##### 2. Specifying the Model

```{r}
# CHANGING THE METHOD TO LINEAR MODEL
## INSERT YOUR CODE HERE ##
```

##### 3. Using Color to Group

```{r}
# INCLUDING COLOR AS A GROUPING VARIABLE
## INSERT YOUR CODE HERE ##
```

##### 4. Combining all the above!

```{r}
## INSERT YOUR CODE HERE ##
```
